---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

```{r, echo=F, include=FALSE}
library(knitr)
knit_child(here::here("Reports", "Results.Rmd"))
```

```{r, echo=F, include=FALSE}
coefCI=function(df, filter_statement, my_term, si){
  my_df=df%>%
    filter(eval(parse(text = filter_statement)))%>%
    unnest(tidied)%>%
    mutate(val=paste0(round(estimate, si), ' (95%CI: ', 
                      round(conf.low, si), ', ',
                      round(conf.high, si), ')'))%>%
    filter(term==my_term)%>%
    pull(val)
  return(my_df)
}
```

# Results


## Study sample descriptive statistics

Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique individuals with DNA methylation data measured on the 450K Illumina array, `r m_obs` samples from `r n_total` individuals had data on additional key covariates (Figure 1). `r n_twovisit` individuals had both age 9 and age 15 samples. Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African ancestry and less likely to be from children of Hispanic ancestry (Supplemental Table 1). Twenty percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alchol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic ancestry.  

We calculated several common summary measures of DNA methylation, including global methylation, epigenetic clocks (Pediatric and GRIM), and polymethylation scores. We calculated polymethylation scores using regression coefficients for 6074 CpGs made from three different models in Joubert et al.: a regression using newborn cordblood samples controlling for cell-type proportions, a regression using newborn cordblood samples not controlling for cell-type proportions and a regression using peripheral blood samples from older children. We also examined the top probe from Joubert et al., AHRR:cg05575921. Many of the methylation summary measures were correlated with each other (Supplemental Figure 1). Notably, polymethylation scores calculated using coefficients from a cell-type corrected regression as weights did not correlate with estimated cell-type proportions, while scores calculated using coefficients from a regression not corrected for cell-type did. Among the `r n_twovisit` individuals with data from both visits, methylation summary measures from the same individual were correlated between the age 9 and age 15 (Supplemental Figure 2). The correlation was strongest for the polymethylation score calculated using weights from a regression for sustained smoking in newborn cordblood with cell-type control (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001). The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3; mean pediatric clock age 9: `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))` age 15: `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))`, mean GRIM clock age 9 `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`, age 15: `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`. The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3). Based on these results, we focused on the polymethylation score from newborn cordblood with cell-type control, the pediatric clock, AHRR:cg05575921 and global methylation for bivariate and multivariable analyses. 

## Bivariate associations between prenatal maternal smoking and DNA methylation summary measures

Mothers who reported smoking during pregnancy had lower income to poverty ratios and were more likely to report prenatal alcohol use, prenatal drug use and postnatal smoking than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European ancestry and had higher prenatal maternal smoking polymethylation scores at both the age 9 and age 15 visit (Table 1). Children of mothers who reported prenatal maternal smoking did not differ from those whose mothers did not in terms of global methylation, epigenetic age estimaated from pediatric clock or estimate cell-type proportions. Compared to children of mothers not reporting prenatal materna smoking, exposed children were significantly more hypomethylated at AHRR:cg05575921 at age 9 (mean methylation `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`) but not age 15 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)` *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`). 

## Multivariable models 

The association between prenatal maternal smoking and the cell-type controlled polymethylation score was also observed in multivariable models controlling for precision and confounder variables (Figure 3; Supplemental Table 2). In the entire sample, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 9 and `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 15. A consistent association was observed when stratifying to only African ancestry children (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`). The direction and magnitude of the association was similar in European and Hispanic ancestry children at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2). Prenatal maternal smoking was associated with hypomethylation at AHRR:cg055975921 at age 9 but not age 15 and not associated with global methylation or epigenetic age from the pediatric clock at either age. Results were similar in linear mixed effect models and in sensitivity models controlling for surrogate variables instead of known covariates (Supplemental Tables 3-4; Supplemental Figure 4-5). 

## Accuracy of polymethylation scores as a biomarker of prenatal maternal smoking 

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A). The polymethylation score with cell-type control had the largest area under the curve (AUC) both when used as a single predictor (age 9 AUC; `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)` ) and when used in models including child sex, maternal income-poverty ratio, immune cell proportion and batch (age 9 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)`). The AUC for the polymethylation score was significantly larger than using only the top CpG from Joubert et al., AHRR:cg05575921 (age 9 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(4)%>%round(3)` , age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(4)%>%round(3)`). Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B). 


